Model Checking

### 2024-01-15 owen/Write Ups/04_model_checking.qmd
working on updating all of this, putting in all the fits and starting some of the other things we discussed in the last meeting with James

We fit this model also to 10 players rated between 2000 and 2200. 
With the same restrictions on the games this gives us 285k games.
We have between 1200 and 89k games for each of these 10 players.

<!-- Down to here in terms of updating this -->

fit2 <- readRDS(here("results/", "lichess2000-2200/", 
                    "all_rated_bullet_model.RDS"))

@@

@@

```{r get-player-names}
#| cache: true

small_data <- lichess_data %>% 
  # filter(Event == "Rated Bullet game") %>% 
  # filter(TimeControl == "60+0") %>% 
  filter(Variant == "Standard") %>% 
  filter(grepl("Rated Bullet game", Event)) #%>% 
  # filter(!grepl("Classical|Correspondence", Event))

users_20_22 <- unique(small_data$Username)

rm(lichess_data)
rm(small_data)

@@

```{r}
names(players) <- paste0("alpha[", 1:length(users_20_22), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit2$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

- Here the effect of a difference in ranking appears to be slightly larger.
This leads to a bigger change in the win probability when the difference
in ranking between the two players is larger.

- The posterior standard deviations of the $\mu$ parameters is slightly
larger in this model, but this could be explained by the smaller number of 
games.

- A smaller sample, but half of these posterior distributions 
of the winner effects contain 0.

@@

We look at 8 players rated between 2300 and 2500. This gives us 160k games,
with players playing 4k-58k games.

```{r}
fit3 <- readRDS(here("results/", "lichess2300-2500/", 
                    "all_rated_bullet_model.RDS"))

mcmc_hist(fit2$draws(variables = c("mu1", "mu2", "gamma1", "gamma2"))) +
  labs(title = "Global Parameters",
       subtitle = "mu1 is Winner Effect, mu2 is Player Effect, Gamma1 is colour,
       Gamma2 is ELO Difference")
```

```{r get users 2300-2500}
#| cache: true

data_path <- here("box_data/lichess2300-2500/")

files <- list.files(data_path)
files <- files[1:8]

small_data <- lichess_data %>% 
  # filter(Event == "Rated Bullet game") %>% 
  # filter(TimeControl == "60+0") %>% 
  filter(Variant == "Standard") %>% 
  filter(grepl("Rated Bullet game", Event)) #%>% 
  # filter(!grepl("Classical|Correspondence", Event))

users_23_25 <- unique(small_data$Username)

rm(lichess_data)
rm(small_data)

@@

## Other Questions

>>>>

TO BE COMPLETED

We fit this model also to all players rated between 2000 and 2200. 
With the same restrictions on the games this gives us 834k games.
We have between 1200 and 89k games for each of these players.

```{r}
#| eval: false

fit2 <- readRDS(here("results/", "Full_Fits", "lichess2000-2200/", 
                    "all_rated_bullet_model.RDS"))

@@

users_20_22 <- readRDS(here("results/", "Full_Fits",
                      "lichess2000-2200/",
                      "users_bullet.RDS"))

@@

```{r}
#| eval: false

@@

```{r}
#| eval: false

names(players) <- paste0("alpha[", 1:length(users_20_22), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit2$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

<!-- - Here the effect of a difference in ranking appears to be slightly larger. -->
<!-- This leads to a bigger change in the win probability when the difference -->
<!-- in ranking between the two players is larger. -->

<!-- - The posterior standard deviations of the $\mu$ parameters is slightly -->
<!-- larger in this model, but this could be explained by the smaller number of  -->
<!-- games. -->

<!-- - A smaller sample, but half of these posterior distributions  -->
<!-- of the winner effects contain 0. -->

@@

We look at all players rated between 2300 and 2500. This gives us 703k games,
with players playing 400-74k games.

```{r}
fit3 <- readRDS(here("results/", "Full_Fits", "lichess2300-2500/", 
                    "all_rated_bullet_model.RDS"))

users_23_25 <- readRDS(here("results/", "Full_Fits",
                      "lichess2300-2500/",
                      "users_bullet.RDS"))

mcmc_hist(fit3$draws(variables = c("mu1", "mu2", "gamma1", "gamma2"))) +
  labs(title = "Global Parameters",
       subtitle = "mu1 is Winner Effect, mu2 is Player Effect, Gamma1 is colour,
       Gamma2 is ELO Difference")
```

@@

## Repeating for Blitz Games

### 1700-1900

fit <- readRDS(here("results/", "Full_Fits", "lichess1700-1900/", 
                    "all_rated_blitz_model.RDS"))

## winner effects
players <- users
names(players) <- paste0("beta[", 1:length(users), "]")

mcmc_hist(fit$draws("beta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 2000-2200

TO BE COMPLETED

```{r}
#| eval: false

fit2 <- readRDS(here("results/", "Full_Fits", "lichess2000-2200/", 
                    "all_rated_blitz_model.RDS"))

## winner effects
players <- users
names(players) <- paste0("delta[", 1:length(users), "]")

mcmc_hist(fit2$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit2$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")

### 2300-2500

TO BE COMPLETED

```{r}
#| eval: false

fit3 <- readRDS(here("results/", "Full_Fits", "lichess2300-2500/", 
                    "all_rated_blitz_model.RDS"))

## Alternative Model

The alternative model is to use only the previous game as history, 
ignoring all other games. This is perhaps more susceptible to 
issues with correlation, which we will need to discuss further.

### 1700-1900 Bullet

fit <- readRDS(here("results/", "Full_Fits", "lichess1700-1900/", 
                    "all_rated_bullet_model_prev.RDS"))

## winner effects
players <- users
names(players) <- paste0("delta[", 1:length(users), "]")

mcmc_hist(fit$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 2000-2200 Bullet

```{r}
fit2 <- readRDS(here("results/", "Full_Fits", "lichess2000-2200/", 
                    "all_rated_bullet_model_prev.RDS"))

fit2$summary(variables = c("mu1", "mu2", "gamma1", "gamma2"))

## winner effects
players <- users_20_22
names(players) <- paste0("delta[", 1:length(users_20_22), "]")

mcmc_hist(fit2$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users_20_22), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit2$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 2300-2500 Bullet

```{r}
fit3 <- readRDS(here("results/", "Full_Fits", "lichess2300-2500/", 
                    "all_rated_bullet_model_prev.RDS"))

mcmc_hist(fit3$draws(variables = c("mu1", "mu2", "gamma1", "gamma2"))) +
  labs(title = "Global Parameters",
       subtitle = "mu1 is Winner Effect, mu2 is Player Effect, Gamma1 is colour,
       Gamma2 is ELO Difference")

fit3$summary(variables = c("mu1", "mu2", "gamma1", "gamma2"))

## winner effects
players <- users_23_25
names(players) <- paste0("delta[", 1:length(users_23_25), "]")

mcmc_hist(fit3$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users_23_25), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit3$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 1700-1900 Blitz

```{r}
fit <- readRDS(here("results/", "Full_Fits", "lichess1700-1900/", 
                    "all_rated_blitz_model_prev.RDS"))

## winner effects
players <- users
names(players) <- paste0("delta[", 1:length(users), "]")

mcmc_hist(fit$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 2000-2200 Blitz

```{r}
fit2 <- readRDS(here("results/", "Full_Fits", "lichess2000-2200/", 
                    "all_rated_blitz_model_prev.RDS"))

fit2$summary(variables = c("mu1", "mu2", "gamma1", "gamma2"))

## winner effects
players <- users_20_22
names(players) <- paste0("delta[", 1:length(users_20_22), "]")

mcmc_hist(fit2$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users_20_22), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit2$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

### 2300-2500 Blitz

```{r}
fit3 <- readRDS(here("results/", "Full_Fits", "lichess2300-2500/", 
                    "all_rated_blitz_model_prev.RDS"))

mcmc_hist(fit3$draws(variables = c("mu1", "mu2", "gamma1", "gamma2"))) +
  labs(title = "Global Parameters",
       subtitle = "mu1 is Winner Effect, mu2 is Player Effect, Gamma1 is colour,
       Gamma2 is ELO Difference")

fit3$summary(variables = c("mu1", "mu2", "gamma1", "gamma2"))

## winner effects
players <- users_23_25
names(players) <- paste0("delta[", 1:length(users_23_25), "]")

mcmc_hist(fit3$draws("delta"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Player Winner Effects")

## indiv effects
names(players) <- paste0("alpha[", 1:length(users_23_25), "]")
player_labels <- as_labeller(players)
mcmc_hist(fit3$draws("alpha"),
          facet_args = list(labeller = player_labels)) +
  labs(title = "Individual Player Effects")
```

## Other Questions of Interest

- How much do player ratings change across their games?

We can look at this for a sample of players, from the 2300-2500 category.

```{r}
## load in the data for this

path_id <- 3

all_data_path <- rep(NA, 3)
all_save_path <- rep(NA, 3)
all_data_path[1] <- here("box_data/lichess1700-1900/")
all_data_path[2] <- here("box_data/lichess2000-2200/")
all_data_path[3] <- here("box_data/lichess2300-2500/")

data_path <- all_data_path[path_id]
files <- list.files(data_path)

small_data <- lichess_data %>%
  # filter(Event == "Rated Bullet game") %>%
  # filter(TimeControl == "60+0") %>%
  filter(Variant == "Standard") %>%
  filter(grepl("Rated Bullet game", Event))

users <- unique(small_data$Username)

small_data$dttm <- as.POSIXct(paste(small_data$UTCDate, small_data$UTCTime),
                              format="%Y-%m-%d %H:%M:%S", tz = "UTC")

small_data %>% 
  mutate(UserELO = ifelse(Username == White, WhiteElo, BlackElo)) %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime, .by_group = TRUE) %>% 
  select(Username, dttm, UserELO) %>% 
  filter(Username %in% users[1:16]) %>% 
  # slice_sample(n = 100) %>% 
  ggplot(aes(dttm, UserELO, colour = as.factor(Username))) +
  geom_line() +
  labs(colour = "User", x = "Date") +
  geom_hline(yintercept = c(2300, 2500), 
             col = "red", linetype = "dashed", alpha = 0.5) +
  facet_wrap(~Username, scales = "free")

Each player by default starts with a rating of 1500.
Note that there is a lot of variability in how much time (and how many games)
each user has played at the "Ratings Bin" they are assigned to. Given
that current opponents are based only on your current rating, this 
should likely be accounted for. A player who has played many 
games where their rating is above 2300 should be different 
to a player playing their first game after attaining that
rating. How do we account for this in the model?

```{r}
## look at different bin
path_id <- 1
data_path <- all_data_path[path_id]
files <- list.files(data_path)

small_data_1 <- lichess_data %>%
  # filter(Event == "Rated Bullet game") %>%
  # filter(TimeControl == "60+0") %>%
  filter(Variant == "Standard") %>%
  filter(grepl("Rated Bullet game", Event))

users_1 <- unique(small_data_1$Username)

small_data_1$dttm <- as.POSIXct(paste(small_data_1$UTCDate,
                                      small_data_1$UTCTime),
                              format="%Y-%m-%d %H:%M:%S", tz = "UTC")

small_data_1 %>% 
  mutate(UserELO = ifelse(Username == White, WhiteElo, BlackElo)) %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime, .by_group = TRUE) %>% 
  select(Username, dttm, UserELO) %>% 
  filter(Username %in% users_1[1:16]) %>% 
  # slice_sample(n = 100) %>% 
  ggplot(aes(dttm, UserELO, colour = as.factor(Username))) +
  geom_line() +
  labs(colour = "User", x = "Date") +
  geom_hline(yintercept = c(1700, 1900), 
             col = "red", linetype = "dashed", alpha = 0.5) +
  facet_wrap(~Username, scales = "free")
```

- How could we control for the experience of a player here? Would this
impact the presence of these winner effects?

## Old Questions
